---
title: "Pipeline"
output:
  html_document:
    fig_width: 9
    fig_height: 8
    fig_caption: yes
    code_folding: hide
  pdf_document: default
---

```{r, echo=FALSE}
knitr::opts_chunk$set( echo = TRUE, code_folding='hide')   
```

## Load simulation data

```{r echo=TRUE}
library(FFT)
library(PTA)
library(leiden) 
cluLeiden = NULL
load("../data/simulation_res.RData")
```

## Analysis pipeline {.tabset}
 
### 1). PCA
1. Users should use normalized data as input for PCA analysis pcaL1().
2. Aooly getPCAN() function to automatically select top principle components (PCs): 
   (1) Compute the derivative of the standard deviations of the principal components.
   (2) Apply linear regression on the PC number and the derivative. Compute the squared error of the predicted derivative and input derivative. 
   (3) Compute the median value of the top 100 values in the predicted squared error.  
   (4) Use the median value as the mean and standard variance of normal distribution. We use right tailed test (p-value > 1 - 0.05) to select the top PCs based on the predicted squared error. 
```{r echo=TRUE}
if(is.null(pcd)){
  pcd = pcaL1(data=daC, pvalue=0.05, pcaSel = 1,  sel=2, center=TRUE, scale=FALSE, ith=1e-2) 
}   

if(is.null(Z)){
  nc = getPCAN(pcd$pca_out,pvalue=0.05, sel=2, ith=1e-2)
  Z = pcd$pca_out$x[,1:nc] 
} 
dim(Z) 
``` 

### 2). UMAP
Compute UMAP using the selected PCs.
```{r echo=TRUE }
library(plotly)
# For plot 
set.seed(1)
if(is.null(uY)){
  emb2 <-  umap::umap(Z)
  uY = emb2$layout
}

### visualize the cell annotations in UMAP
plotAlls(X=uY,draw=1, t=NULL, cluid=as.character(branch),
         type=2,size =6,opacity=1, titlename=paste0("UMAP: PC number = ", ncol(Z)))

```  

### 3). Clustering
1. Using different k size for building corresponding kNN graph for clustering with cmcluster(). 
   For each kNN graph, apply community detection (Leiden or Louvain) algorithm for clustering.
```{r echo=TRUE }
ks = seq(4,20, by=2)
if(is.null(cluLeiden)){
  tclu3 = proc.time()
  cluLeiden = list()
  #setLeignPath()
  for(i in 1:length(ks)){
    cluLeiden[[i]] = cmcluster(Z, ksize = ks[i], cmmethod="leiden")
  }
  tclu4 = proc.time() - tclu3  
}
``` 

2. User can visualize the clustering results in UMAP and select a proper clustering results.

```{r echo=TRUE }
lp <- htmltools::tagList()
for (i in 1:length(ks)) { 
  cluid = cluLeiden[[i]]$cluid
  table(cluid)
  p1 = plotAlls(X=uY,draw=1, t=NULL, cluid=as.character(cluid), type=2,size =6,opacity=1,
         titlename=paste("UMAP,k=", ks[i], sep=""))
  lp[[i]] <- as.widget(p1)
}
lp
``` 

### 4). Select a proper clustering
  
```{r echo=TRUE}
k=5
cluid = cluLeiden[[k]]$cluid
table(cluid)
plotAlls(X=uY,draw=1, t=NULL, cluid=as.character(cluid), type=2,size =6,opacity=1,
         titlename=paste("UMAP,k=", ks[k], sep=""))

library("netbiov")
## visualize the kNN graph
plotCommNet(Z, ksize=ks[k], cluid, colname= rownames( brewer.pal.info)[12],
            gtype = "Kamada-Kawai", seedN = 1)

``` 

### 5). Compuate L-ISOMAP
1. Based on selected clustering, we build kNN graph from corresponding k size.
2. For each cluster, select the the cell which has most connections in kNN graph as landmark points use searchPeaks().
3. Build 3D L-ISOMAP based landmark points and kNN graph with autoLISOMAP().
```{r echo=TRUE}  
centid = getCluCenterID(Z, cluid)
peakid = searchPeaks(gg=cluLeiden[[k]]$knng$g, cluid)
X  <- autoLISOMAP(Z, peaks=peakid, ndim = 3, L1=NULL, L2=NULL, N=1)   
``` 

Visualize cell annotation in 3D L-ISOMAP 

```{r echo=TRUE}  
PZ = rbind(uY, uY[peakid,])
pc = c(cluid, rep("peaks", length(peakid)))
plotAlls(X=PZ,draw=1, t=NULL, cluid=as.character(pc), type=2,size =6,opacity=1,
         titlename=paste("Peaks in UMAP,k=", ks[k], sep="")) 
#### 4. Visualization
#plotAlls(X=X,draw=3, t=NULL, cluid=as.character(branch), type=2,size =6,opacity=1, titlename="ISOMAP")
plotAlls(X=X,draw=3, t=NULL, cluid=as.character(cluid), type=2,size =5,opacity=1, titlename="ISOMAP") 
``` 

### 6). Trajectory anlysis 
1. Calculate the neighbor distance matrix and graph distance matrix for the clusters based kNN graph.
2. User can specify the root cluster and leaf clusters to build a spanning tree.
   User can organize the leaf structure by list format:
   For example: leaves=list(a=c(5,7,"parallel"), b=c(8,9,"parallel"), c=c(16,12,"linear"), d=10, f=4)
   Here, cluster 5 and 7 contain cells in branch 9 and 10, and they has same parent. Hence, they are parallel. 
   cluster 8 and 9 contain cells in branch 6 and 5, and they has same parent. Hence, they are parallel.
   cluster 12 and 16 contain cells from same branch 12. We set them a linear connected.
   cluster 10 and 4 contain cells in branch 11 and 8. 
   
```{r echo=TRUE}
##1. Compute neighbor distance of clusters in graph built by PCA  
msM = getCluNeighbors(Z, k=ks[k], cluid=cluid)
##2. Compute graph distance matrix
dsM = distGraphClu(Z, ksize=ks[k], cluid=cluid) 
``` 


```{r echo=TRUE}
rt = 2 # set root cluster
g1 = getMSTLeafRootISO_3(neigDist = msM, cenDist=dsM, cluid=cluid, L2=3, root=rt,
                         leaves=list(a=c(5,7,"parallel"), b=c(8,9,"parallel"),
                                     c=c(16,12,"linear"), d=10, f=4))
``` 

### 7). Trajectory visualization 
1. User can visualize the trajectory by pie tree plot
```{r echo=TRUE}
clucells = getCluCellTh(cluid=cluid, branch, th=0.2, subN=10)
g2 <- set.vertex.attribute(g1$g5, "name", value=clucells)
ctyM = getCellSM(cluid, branch)
cellN =  as.character(names(table(branch)))

g3 <- set.vertex.attribute(g1$g5, "name", value= paste("Y_",1:max(cluid),sep=""))
pieTreeRoot(mst=g3, root=rt, ctyM, cellN, vs=1, ew=1, cex=2, seed=1,colorname="Paired") # Plot the cluster tree 
``` 

2. User can visualize the trajectory in 3D L-ISOMAP
```{r echo=TRUE}
#### Plot 3D trajectory
plot3DTreeCells(X[,1:3], peaks =NULL, cluid= as.character(cluid), celltype = as.character(branch),
                colorname="Paired", mst=g1$g5, pointsize = 6,labelsize=15, lineW=10)
#plot3DTreeCells(X[,1:3], peaks =NULL, cluid= as.character(cluid),colorname="Paired",
#                celltype = as.character(cluid),mst=g1$g5, pointsize = 6,labelsize=15, lineW=10)

```

3. User can visualize the trajectory by cell cluster tree
```{r echo=TRUE} 
####   Plot the tree
Y <- getCenters(X, cluid)
cluidS = as.character(cluid)
anno =  cbind(cluidS, as.character(branch))
colnames(anno) = c("cluster", "celltype")
anno = as.data.frame(anno)

g3 <- set.vertex.attribute(g1$g5, "name", value= paste("Y_",1:max(cluid),sep=""))
#g3 <- set.vertex.attribute(g1, "name", value= paste("Y_",1:max(cluid),sep=""))
plotCellTree(X=X, uY, cluid=cluid, mst=g3, root=rt, colorby="celltype", anno=anno, height=0.2,
             width=0.3, pointSize=2, edgeW=0.5, textSize=6, legendTextS=20, legendColS=10, hv=0.5,
             vv=-1.2, colorName = "Paired")
```

### 8). Trajectory pseudo time 
1. User can compute pseudo time based on the trajectory.
```{r echo=TRUE}
rt = 2# root
g2 <- set.vertex.attribute(g1$g5, "name", value= paste("Y_", 1:max(cluid),sep = ""))
res = simpPTime(Z, cluid=cluid,  cenTree=g2,  root=rt)

# save(a=pcd, b=Z, c=uY, d=cluLeiden, e=X, f=g1, g=branch, h=daC, i=Y, j=res,k=g2, file = "../data/simulation_res.RData")

```

2. User can visualize the pseudo time in UMAP.

```{r echo=TRUE}
plotAlls(X=uY,draw=1, t=res$pt, cluid=as.character(branch), type=1,size =6,opacity=1, titlename="UMAP")
#plotAlls(X=X,draw=3, t=res$pt, cluid=as.character(branch), type=1,size =6,opacity=1, titlename="ISOMAP")

```


### 9). PTA analysis for each lineage 
User can apply PTA for each lineage
1. User should specify each lineage first as input of PTA
2. To reduce the computation time, user can do average on the cells along the lineage

```{r echo=TRUE}
lpath1 = c(2,15,11,9)
lpath2 = c(2,15,11,8)
lpath3 = c(2,14,13,6,4)
lpath4 = c(2,14,13,6,3,7)
lpath5 = c(2,14,13,6,3,5)
lpath6 = c(2,14,13,1,16,12)
lpath7 = c(2,14,13,1,10)
cluid = cluLeiden[[5]]$cluid
fpath = "../data/simu_PTA_lineage_1.RData"

if(!file.exists(fpath)){ 
  usePTA(sd=daC, cluid=cluid, path=lpath1, pt = res$pt, fname=fpath,
       it = 25, intv = 200, feature.flag = TRUE,
       err_flag = 1, scale=FALSE, nCores=1) 
} 
load(fpath)
```

### 10). PTA Visulization in rank 1 in lineage c(2,15,11,9)
 
```{r echo=TRUE}
library(reshape2)
library(colorRamps)
library(grid)
library(gridExtra)
library(pheatmap)
singlePath = c(2,15,11,9)
cid = singlePath
da = NULL
for(i in 1:length(cid)){
  k= which(cluid==cid[i])
  pt = res$pt[k]
  od = order(pt)
  k = k[od]
  da = cbind(da, daC[,k]) 
} 

plotHeatMap <- function(da1,tv, fz=20, rowsize = 6, colsize=4,barsize=10, showRname=F, showCname=F,
                        colname=NULL, title=""){#
  #cs = colorRampPalette(rev(c("#D73027", "#FC8D59", "#FEE090",
  #                                                 "#FFFFBF", "#E0F3F8", "#91BFDB", "#4575B4")))(100);
  if(is.null(colname)){
    #cs = colorRampPalette(rev(c("#D73027", "#FC8D59", "#FEE090",
    #                            "#FFFFBF", "#E0F3F8", "#91BFDB", "#4575B4")))(100);
    cs = colorRampPalette(c("#0392cf","#7bc043", "#fdf498","#ffff66","#ff3300"))(50)
  }else{#colname="Spectral"
    cs = colorRampPalette(brewer.pal(brewer.pal.info[colname,1],colname))(100)
  } 
  
  a = pheatmap(da1, color = cs, breaks = NA, scale = "none", cluster_rows = F,
               cluster_cols = F,   legend = TRUE, annotation = NA,
               annotation_colors = NA, annotation_legend = TRUE, filename = NA, width = NA, 
               height = NA, main = title, 
               show_rownames=showRname,show_colnames=showCname,
               silent=TRUE, fontsize = fz,fontsize_row = rowsize, fontsize_col = colsize)
  
  if(tv[1] < 0){
    od = order(tv, decreasing = TRUE)
    tv = tv[od]
  }else{
    od = order(tv)
    tv = tv[od]
  } 
  df <- data.frame(top_v=1:length(tv), loadings=tv)
  p<- ggplot(data=df, aes(x=top_v, y=loadings)) +
    geom_bar(stat="identity")  + coord_flip() +
    theme(panel.background = element_blank(), axis.text=element_text(size=barsize),
          axis.title=element_text(size=barsize,face="bold"),
          axis.ticks.x=element_blank())+  
    scale_x_continuous(position = "top") +
    scale_y_continuous(position = "right")
  #+ theme_classic()
  #+ theme_minimal()
  lm=rbind(c(1,1,1,1,1,2))
  grid.arrange(grobs = list(a[[4]],p), layout_matrix = lm) 
  
}

```

1.Plot the main trends

```{r echo=TRUE}
str="Main trends in simulation lineage 2->9" 
feature.flag=TRUE

titlefontSize = 2

# out$uName
##################all trends
mout = out
timevec = 1:nrow(mout$B)
S1 = mout$B %*% mout$beta
f1 <- splinefun(timevec, S1)
T = length(timevec)
max1 = max(mout$prd) 
min1 = min(mout$prd)

mout = out$rank2
S2 = mout$B %*% mout$beta
f2 <- splinefun(timevec, S2)
max2 =  max(mout$prd) 
min2 =  min(mout$prd) 

mout = out$rank3
S3 = mout$B %*% mout$beta
f3 <- splinefun(timevec, S3)
max3 =  max(mout$prd) 
min3 =  min(mout$prd) 

ymin = min(c(S1,S2,S3))
ymax = max(c(S1,S2,S3))

vmax = max(max1, max2, max3)
vmin = min(min1, min2, min3) 

if(1==1){
  plot(x= timevec, y= S1, ylim=c(ymin, ymax), type="p", xlab="Time", ylab="S", lwd=2, font=2, col="black", main= str, cex.main= titlefontSize)
curve(f1(x), from=timevec[1], to=timevec[T], col="red", lty=1,lwd=2, add=TRUE)

points(x= timevec, y= S2, type="p", xlab="Time", ylab="S", lwd=2, font=2, col="black")
curve(f2(x), from=timevec[1], to=timevec[T], col="blue", lty=1,lwd=2, add=TRUE)

points(x= timevec, y= S3, type="p", xlab="Time", ylab="S", lwd=2, font=2, col="black")
curve(f3(x), from=timevec[1], to=timevec[T], col="green", lty=1,lwd=2, add=TRUE)
par(xpd = T)
legend(1, ymax+1, title = " ",  legend=c('Rank1','Rank2','Rank3'), 
       col =  c('red','blue','green'), horiz = F,bty='n',
       pch = 14,pt.cex = 2,  cex=1.5,ncol= 10 ) 
}
  
plot(x= timevec, y= S1, ylim=c(ymin, ymax), type="p", xlab="Time", ylab="S", lwd=2, font=2, col="black", main= str, cex.main= titlefontSize)
curve(f1(x), from=timevec[1], to=timevec[T], col="red", lty=1,lwd=2, add=TRUE)

points(x= timevec, y= S2, type="p", xlab="Time", ylab="S", lwd=2, font=2, col="black")
curve(f2(x), from=timevec[1], to=timevec[T], col="blue", lty=1,lwd=2, add=TRUE)
par(xpd = T)
legend(1, ymax+4, title = " ",  legend=c('Rank1','Rank2' ), 
       col =  c('red','blue'), horiz = F,bty='n',
       pch = 16,pt.cex = 2,  cex=1.5,ncol= 10 ) 


PTA.plotCV(err,feature.flag = feature.flag)
 
plot(mout$a[,1],type="p", xlab="genes", ylab="a", lwd=2, font=2)
 
TH1 = 0.03
corTH = 0 
mout = out
na <- c(length(mout$a[mout$a>TH1]), length(mout$a[mout$a<-TH1]),length(mout$a[abs(mout$a)<TH1]) ) 
names(na) = c(paste('a>',TH1),
              paste('a<',-TH1),
              paste('|a|<',TH1) )
na

gid = 1:length(mout$a)
uid = gid[mout$a>TH1]
did = gid[mout$a<-TH1]
 
# up and down expressed marker genes 
upgene1 = as.character(out$uName[uid])
downgene1 = as.character(out$uName[did]) 

############################ Plot Trend
timevec = 1:nrow(mout$B)
S = mout$B %*% mout$beta
f1 <- splinefun(timevec, S)
T = length(timevec)
plot(x= timevec, y= S, type="p", xlab="Time", ylab="S", lwd=2, font=2, col="black",ylim=range(S), main="Trend")
curve(f1(x), from=timevec[1], to=timevec[T], col="red", lty=1,lwd=2, add=TRUE)
```

2.Plot gene expression heatmap ranked by PTA score.

```{r echo=TRUE }
################################ predicted marker
mout = out
rownames(mout$prd) = mout$uName 
nid = which( abs(mout$a[,1]) > TH1)  
uWt = mout$a[nid,1] 
od = order(uWt,decreasing = TRUE) 
uid = nid[od] 
da1 = mout$prd[uid,]   
plotHeatMap(da1, uWt[od], fz=20, title = "Predicted data gene rank 1") 
 
x = mout$x[,1,]
rownames(x) = mout$uName 
nid = which( abs(mout$a[,1]) > TH1)  
uWt = mout$a[nid,1] 
od = order(uWt,decreasing = TRUE) 
uid = nid[od] 
da1 = x[uid,]   

cuWt = uWt[od]
corS= apply(da1, 1, function(x) {cor(x, S1) }) 
id  = which(abs(corS)>=corTH) 

plotHeatMap(da1[id,], cuWt[id], fz=20, title = "Raw data gene rank 1") 
   
kid = getMapId(rownames(da), rownames(da1[id,]))
plotHeatMap(da[kid,], cuWt[id], fz=20, showRname=T, title = "Raw data gene rank 1") 

```  


### 11). PTA Visulization in rank 2 in lineage c(2,15,11,9)
1.Plot the main trends
```{r echo=TRUE }
TH1 = 0.03
mout = out$rank2

############################ plot the main trends
PTA.plotCV(out$err2,feature.flag = feature.flag)
 
plot(mout$a[,1],type="p", xlab="genes", ylab="a", lwd=2, font=titlefontSize)
 
na <- c(length(mout$a[mout$a>TH1]), length(mout$a[mout$a<-TH1]),length(mout$a[abs(mout$a)<TH1]) ) 
names(na) = c(paste('a>',TH1),
              paste('a<',-TH1),
              paste('|a|<',TH1) )
na


gid = 1:length(mout$a)
uid = gid[mout$a>TH1]
did = gid[mout$a<-TH1]
 
# up and down expressed marker genes 
upgene2 = as.character(out$uName[uid])
downgene2 = as.character(out$uName[did]) 

############################ Plot Trend
timevec = 1:nrow(mout$B)
S = mout$B %*% mout$beta
f1 <- splinefun(timevec, S)
T = length(timevec)
plot(x= timevec, y= S, type="p", xlab="Time", ylab="S", lwd=2, font=titlefontSize, col="black",ylim=range(S), main="Trend")
curve(f1(x), from=timevec[1], to=timevec[T], col="red", lty=1,lwd=2, add=TRUE)


``` 


2.Plot gene expression heatmap ranked by PTA score.

```{r echo=TRUE }
################################ predicted marker
mout = out$rank2
rownames(mout$prd) = out$uName 
nid = which( abs(mout$a[,1]) >TH1)  
uWt = mout$a[nid,1] 
od = order(uWt,decreasing = TRUE) 
uid = nid[od] 
da1 = mout$prd[uid,]   
plotHeatMap(da1, uWt[od], fz=20, title = "Predicted data gene rank 2") 
 
x = out$x[,1,]
rownames(x) = out$uName 
nid = which( abs(mout$a[,1]) >TH1)   
uWt = mout$a[nid,1] 
od = order(uWt,decreasing = TRUE) 
uid = nid[od] 
da1 = x[uid,]   
 
corS= apply(da1, 1, function(x) {cor(x, S2) }) 
id  = which(abs(corS)>=corTH)
cuWt = uWt[od]

plotHeatMap(da1[id,], cuWt[id], fz=20, title = "Raw data gene rank 2") 

kid = getMapId(rownames(da), rownames(da1[id,]))
plotHeatMap(da[kid,], cuWt[id], fz=20, showRname=T, title = "Raw data gene rank 2") 

``` 

### 12). PTA Visulization in rank 3 in lineage c(2,15,11,9)

1.Plot the main trends
```{r echo=TRUE }
TH1 = 0.03
mout = out$rank3

############################ plot the main trends
PTA.plotCV(out$err3,feature.flag = feature.flag)
 
plot(mout$a[,1],type="p", xlab="genes", ylab="a", lwd=2, font=titlefontSize)
 
na <- c(length(mout$a[mout$a>TH1]), length(mout$a[mout$a<-TH1]),length(mout$a[abs(mout$a)<TH1]) ) 
names(na) = c(paste('a>',TH1),
              paste('a<',-TH1),
              paste('|a|<',TH1) )
na


gid = 1:length(mout$a)
uid = gid[mout$a>TH1]
did = gid[mout$a<-TH1]
 
# up and down expressed marker genes 
upgene2 = as.character(out$uName[uid])
downgene2 = as.character(out$uName[did]) 

############################ Plot Trend
timevec = 1:nrow(mout$B)
S = mout$B %*% mout$beta
f1 <- splinefun(timevec, S)
T = length(timevec)
plot(x= timevec, y= S, type="p", xlab="Time", ylab="S", lwd=2, font=titlefontSize, col="black",ylim=range(S), main="Trend")
curve(f1(x), from=timevec[1], to=timevec[T], col="red", lty=1,lwd=2, add=TRUE)


``` 


2.Plot gene expression heatmap ranked by PTA score.

```{r echo=TRUE }
################################ predicted marker
mout = out$rank3
rownames(mout$prd) = out$uName 
nid = which( abs(mout$a[,1]) >TH1)  
uWt = mout$a[nid,1] 
od = order(uWt,decreasing = TRUE) 
uid = nid[od] 
da1 = mout$prd[uid,]   
plotHeatMap(da1, uWt[od], fz=20, title = "Predicted data gene rank 3") 
 
x = out$x[,1,]
rownames(x) = out$uName 
nid = which( abs(mout$a[,1]) >TH1)   
uWt = mout$a[nid,1] 
od = order(uWt,decreasing = TRUE) 
uid = nid[od] 
da1 = x[uid,]   
 
corS= apply(da1, 1, function(x) {cor(x, S2) }) 
id  = which(abs(corS)>=corTH)
cuWt = uWt[od]

plotHeatMap(da1[id,], cuWt[id], fz=20, title = "Raw data gene rank 3") 

kid = getMapId(rownames(da), rownames(da1[id,]))
plotHeatMap(da[kid,], cuWt[id], fz=20, showRname=T, title = "Raw data gene rank 3") 

``` 
## {-}  
